Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2022-03-30 publications/frontiers/dora_outlier_detection_frontiers.tex
Minor revisions from reviewers. #79

The ability to automatically detect out-of-distribution samples in large data 
sets is of interest for a wide variety of scientific domains. Depending on the
application setting, this capability is also commonly referred to as anomaly
detection, outlier detection, or novelty detection. More broadly, this is 
referred to as out-of-distribution (OOD) detection. In general, the goal of 
OOD detection systems is to identify samples that deviate from the majority
of samples in a dataset in an unsupervised manner~\cite{pimentel2014review}. 
In machine learning, these methods are commonly used for identifying 
mislabeled or otherwise invalid samples in a dataset. When working with 
science datasets, OOD detection can be used for 
cleaning datasets, e.g., flagging ground-truth labels with GPS or human
entry error or identifying wrongly categorized objects in a catalog. 
It could also be used for discovery, e.g., to flag novel samples in order 
to guide instrument acquisition or scientific analysis. Another application
is the detection of rare objects that are known to exist but the known
examples are too few to create a large enough labeled dataset for 
supervised classification algorithms.

Despite wide differences in applications, data types, and dimensionality,
the same underlying machine learning algorithms can be employed across 
all of these domains. A challenge for applying them however is that domain
scientists do not always have the programming or machine learning background
to apply the algorithms themselves using existing tools. Given the widespread 
applicability and transferability of OOD methods, the scientific community 
would benefit from a tool that made it easy for them to apply popular outlier
detection algorithms to their science datasets. We created DORA 
(Domain-agnostic Outlier Ranking Algorithms) 
to provide a tool for applying outlier 
detection algorithms to a variety of scientific data sets with minimal coding
required. Users need only to specify details for their data/application 
including the data type, location, and algorithms to run in an experiment
configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
(DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
Reed-Xiaoli (RX) detector~\citep{reed1990adaptive},  
Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
autoencoder (PAE). 
Each algorithm assigns an outlier 
score to each sample in a given dataset. 
DORA provides results organization and visualization 
modules to help users process the results, including sorting samples by outlier 
score, evaluating outlier recall for a set of known/labeled outliers, clustering 
groups of similar outliers together, and web visualization. 
We demonstrated how DORA
facilitates application, evaluation, and interpretation of outlier detection
methods by performing
experiments for three real-world datasets from Earth science, planetary
science, and astrophysics, as well as one benchmark dataset 
(MNIST/Fashion-MNIST). We found that no single algorithm
performed best across all datasets, underscoring the need for a tool 
that enables comparison of multiple algorithms.

@@

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{hayat2021self} 
as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised}, change~\cite{touati2020anomaly}, 
and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. DORA aims to fill the need 
for tools that facilitate application, evaluation, and interpretation of outlier
detection methods.

@@

% This section was specific to IAAI, can change the name, remove, or keep
% as is if we want.
\subsection{Path to deployment} 
Our goal is for DORA to enable increased application and benefit of
outlier detection methods in real-world scientific use cases. We have 
designed DORA to make it as easy as possible for scientists to apply
algorithms and to compare and interpret their results; users need only
to specify the specifics of their data (e.g., path, data type) in a 
configuration file to start running experiments and seeing results. We will
make DORA publicly available as a pip-installable package to make it
easy to integrate into existing scientific workflows. 
The source code and example datasets used in this study will also be 
open-sourced, enabling DORA to be improved and expanded by the 
machine learning and domain science communities. 
In addition, DORA will be infused into the scientific workflows for the
three use cases we demonstrated results for in this study.
The DORA code is openly accessible via Github at
\url{https://github.com/nasaharvest/dora}.
%the use cases from this will also lead to infusion in real scenarios

>>>>

The ability to automatically detect out-of-distribution samples in large data 
sets is of interest for a wide variety of scientific domains. Depending on the
application setting, this capability is also commonly referred to as anomaly
detection, outlier detection, or novelty detection. More broadly, this is 
referred to as out-of-distribution (OOD) detection. In general, the goal of 
OOD detection systems is to identify samples that deviate from the majority
of samples in a dataset in an unsupervised manner~\cite{pimentel2014review}. 
In machine learning, these methods are commonly used for identifying 
mislabeled or otherwise invalid samples in a dataset 
~\citep{liang2018enhancing,bohm2020probabilistic}. 
%
When working with science datasets, OOD detection can be used for 
cleaning datasets, e.g., flagging ground-truth labels with GPS or human
entry error or identifying wrongly categorized objects in a catalog
~\citep{wagstaff:des-anom20,lochner2021astronomaly}. 
It could also be used for discovery, e.g., to flag novel samples in order 
to guide instrument acquisition or scientific analysis
~\citep{wagstaff:demud13,wagstaff:rover-novelty20,
kerner2020comparison,kerner2020analysis}.
Another application
is the detection of rare objects that are known to exist but the known
examples are too few to create a large enough labeled dataset for 
supervised classification algorithms~\citep{chang2002anomaly,zhou2016novel}.

Despite wide differences in applications, data types, and dimensionality,
the same underlying machine learning algorithms can be employed across 
all of these domains. A challenge for applying them however is that domain
scientists do not always have the programming or machine learning background
to apply the algorithms themselves using existing tools. Given the widespread 
applicability and transferability of OOD methods, the scientific community 
would benefit from a tool that made it easy for them to apply popular outlier
detection algorithms to their science datasets. We created DORA 
(Domain-agnostic Outlier Ranking Algorithms) 
to provide a tool for applying outlier 
detection algorithms to a variety of scientific data sets with minimal coding
required. Users need only to specify details for their data/application 
including the data type, location, and algorithms to run in an experiment
configuration file. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
Isolation Forest, Discovery via Eigenbasis Modeling of Uninteresting Data 
(DEMUD)~\citep{wagstaff:demud13}, principal component analysis (PCA),
Reed-Xiaoli (RX) detector~\citep{reed1990adaptive},  
Local RX, negative sampling~\cite{sipple:neg-sampling20}, and probabilistic 
autoencoder (PAE). 
Each algorithm assigns an outlier 
score to each sample in a given dataset. 
DORA provides results organization and visualization 
modules to help users process the results, including sorting samples by outlier 
score, evaluating outlier recall for a set of known/labeled outliers, clustering 
groups of similar outliers together, and web visualization. 
We demonstrated how DORA
facilitates application, evaluation, and interpretation of outlier detection
methods by performing
experiments for three real-world datasets from Earth science, planetary
science, and astrophysics, as well as one benchmark dataset 
(MNIST/Fashion-MNIST). 
%We found that no single algorithm
%performed best across all datasets, underscoring the need for a tool 
%that enables comparison of multiple algorithms.

The key contributions of this paper are:
\begin{itemize}
\item A new pipeline, DORA, for performing outlier detection experiments using 
several AI algorithms that reduces the effort and expertise required for 
performing experiments and comparing results from multiple algorithms
\item Using experiments for a diverse set of real world datasets and 
application areas, we show that no single algorithm performs best for all 
datasets and use cases, underscoring the need for a tool that compares 
multiple algorithms
\item We provide publicly available code for running and contributing to the 
DORA pipeline and datasets that can be used for reproducing experiments or 
benchmarking outlier detection methods 
\end{itemize}

@@

In astrophysics, 
outlier detection methods have been used to identify astrophysical
objects with unique characteristics~\cite{hayat2021self} 
as well as data or modeling 
artifacts in astronomical 
surveys~\citep{wagstaff:des-anom20,lochner2021astronomaly}. 
Example outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised}, change~\cite{touati2020anomaly}, 
and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020comparison,
wagstaff:demud13}.
%
These examples show the benefit of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the many available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. DORA aims to fill the need 
for tools that facilitate application, evaluation, and interpretation of outlier
detection methods.

@@

% This section was specific to IAAI, can change the name, remove, or keep
% as is if we want.
\subsection{Open code and data} 
Our goal is for DORA to enable increased application and benefit of
outlier detection methods in real-world scientific use cases. We have 
designed DORA to make it as easy as possible for scientists to apply
algorithms and to compare and interpret their results. Users need only
to specify the specifics of their data (e.g., path, data type) in a 
configuration file to start running experiments and seeing results for 
their own datasets and use cases. DORA is
publicly available and can be installed using pip via Github, making it
easy to integrate into existing scientific workflows. 
The datasets used in this study are also publicly available via Zenodo. 
This enables DORA to be improved and expanded by the 
machine learning and domain science communities. 
If a researcher wants to use DORA for a dataset with a type that is not yet
supported, they can contribute a new data loader by creating a subclass
that extends the \texttt{DataLoader} abstract base class. Similarly, new 
results interpretation modules can be added by creating a subclass of the 
\texttt{ResultsOrganization} abstract base class. A new outlier ranking
algorithm can be added by writing a new python module that defines
a subclass of the \texttt{OutlierDetection} abstract base class and implements 
the required functions for scoring and ranking samples, following the 
existing algorithm modules named \texttt{*\_outlier\_detection.py}.
In addition, DORA will be infused into the scientific workflows for the
three use cases we demonstrated results for in this study.
The DORA code can be accessed at
\url{https://github.com/nasaharvest/dora} and datasets 
at \url{https://doi.org/10.5281/zenodo.5941338}.
%the use cases from this will also lead to infusion in real scenarios
